# 13-01-2026
```{r}
2+2
```


## working with packages in r
packages in R
anyone can create an r package, this makes it extensible

```{r}
# package installation

install.package{"tableone"} #installation is needed once
library(tableone) #to use it, load it library`
```


# 14-01-2026 lesson
```{r}
# ebola data analysis
# name
# date
# link - http://bit.ly/view-ebola-data

# load packages
if(!require(pacman)) install.packages("pacman")
pacman::p_load(
    tidyverse, #this is a meta package
    inspectdf,
    plotly, #package for plots 
    janitor,
    visdat,
    esquisse
)
```


# load your data
```{r}
esl <- read_csv("ebola_sierra_leone.csv")
# use relative links if not in same folder

glimpse(esl)


# Why `glimpse()`?
# - Works in Quarto
# - Shows structure + data types
# - Part of `dplyr` (tidyverse)
# useful If you’re just exploring interactively (not rendering)

```


# look at the data
```{r}
esl <- read_csv("ebola_sierra_leone.csv")

```
View(esl)


```{r}
```{r}
install.packages("readr")
library(readr)

esl <- read_csv("ebola_sierra_leone.csv")

```

```{r}
installed.packages()[, "Package"]
```

# data exploration
head(dataset, n=number of rows) #returns first six rows
tails() shows bottom rows 
by default six

ncol(dataset)
nrow(dataset)
dim() dimension x y

summary() 

all functions are from base r packages
?function name to know more about function 

# general overview

visdat::vis_dat(dataset) #shows visualization of teh data
which are characters, date, numeric, NA



```{r}
inspectdf::inspect_cat(esd)
```

# categorical overview
inspect::inspect_cat() send output to showplot function 
this is function nesting

show_plot(inspect_cat(dataset))

ggplotly(
    show_plot(inspect_cat(dataset))
) #shows interactive document

better = show_plot(inspect_cat(dataset))


ggplotly(
    better
)

# numerical 
num = <- show_plot(inspect_num(dataset))

ggplotly on num summary